When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-22 doc/paper/vpd_et_paper.tex
added makefile for creating paper, works great

@@

\begin{figure}[!p]
\centering
\includegraphics[width=20pc]{./fig01.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  a placeholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

@@

By construction, the variability in the $\sigma$ term (Equation \ref{sigma}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be close to 1. Figure \ref{lai_fig} presents the histogram of calculated $\sigma$s with outliers removed (lowest and highest 5\% percent). All remaining $\sigma$ values are close to unity (i.e. $O(1)$), which provides further confidence in the model framework and in the data quality.

\begin{figure}[!p]
\centering
\includegraphics[width=20pc]{./fig02.png}
\caption{Histogram of $\sigma$ values calculated for each site and time according to Equation \ref{sigma}. The lowest and highest 5\% are removed as outliers, as well as any values below 0. The curve is normalized such that its area is 1. }
\label{lai_fig}
\end{figure}

An additional concern is that $\sigma$ may in fact be correlated with $VPD$, in which case the dependence would need to be accounted for when taking the derivative. However, the joint distribution of $\sigma$ and VPD shows that $\sigma$ is very weakly a function of VPD (Figure \ref{lai_vpd_fig}). Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $VPD$.

\begin{figure}[!p]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. \textbf{***note I am redoing/cleaning up this plot, ``lai'' should read sigma***}}
\label{lai_vpd_fig}
\end{figure}

@@

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive CRO and water conservative CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$. It also has a slightly higher uWUE ($2.32$ -units-) compared to CSH ($uWUE=2.11$ -units-). \explain{MAYBE A DUMB QUESTION BUT CAN'T uWUE AND G1 BE ANALYTICALLY RELTED> I THINK AT LESAT YOU SHOULD MAKE CLEAR THAT UWUE IS AT THE ECOYSTEM SCALE (DID YOU FIT IT OR USED PUBLISHED DATA-I AM NOW CONFUSED)}. These differences in slope parameter cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always increases with increasing VPD for the more water intensive CRO, while ET almost always decreases with increasing VPD for the more water conservative CSH (Figure \ref{term3}). CRO evolved or were bred to prioritize GPP and yield and are thus not water conservative. They are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

As expected, the slope parameter ($g_1$) is a primary determinant of the VPD dependence for the sign term shown in Figure \ref{term3}. Plants that are more conservative (small $g_1$) will tend to reduce ET with increasing VPD, and will be very effective at reducing ET, especially at low VPD. However, at very high VPD, gradients in vapor pressure at the leaf scale will be so strong that they will begin to dominate leaf-scale plant strategies of stomatal (partial) closure (parameterized with $g_1$). As a result, ET response will begin to asymptote towards its ecosystem-scale values irregardless of leaf-scale strategy and stomatal regulation (parameterized with $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf stomatal response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize our theoretical insights, Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET with increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD response dependence, while CRO ($g_1 = 183.1$) has the smallest VPD dependence. ENF are less willing to trade water for GPP, so the difference between the response at low VPD when leaf response dominates and high VPD when vapor pressure gradients dominate is much greater. CROs on the other hand have a more aggressive water use strategy and are more willing to trade water for photosynthesis, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively small.

Figure \ref{term3} clearly shows, according to our theory, that for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to reduce ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates in typical environmental conditions, emphasizing the water conservative strategy of those plants. For CRO on the other hand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants operate with an aggressive water usage strategy, are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on environmental conditions, with about half of the observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, ET response to VPD is still far smaller than it would be for potential evaporation $\partial PET/\partial VPD$, i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface response to a change in VPD will always me more negative than if the plants are not included in the physics (PET).

\begin{figure}[!p]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***I think I'm going to remove the bottom plot: its discussion didn't fit in well with the text, and the box plots test inclusion of uncertainty***}}
\label{sign_plot}
\end{figure}

@@

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./fig04b.pdf}
\caption{Primary sources of variability for the scaling term, as a function of PFT. Note the change in axis scale for each plot. The 5th-95th percentile range of temperature is plotted at the 5th, 25th, 50th, 75th, and 95th percentiles of aerodynamic conductance, as observed for each PFT's sites.}
\label{scale_vary}
\end{figure}

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
While Section \ref{stats_sec} presented simple statistics of observed response, we now compare more sophisticated distributions of the observed response to our theory presented in Section \ref{sign_term}. Figure \ref{agu_fig} shows observed the distribution of of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

As Section \ref{stats_sec} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, due to some bias in $uWUE$. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics of Section \ref{stats_sec}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance tend to not only observe a single species but many differnent ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{agu_fig}.

@@

While the above discussion shows that our theory has some limitations when applied to some PFTs, especially for crops and grasslands, it does well for DBF and CSH PFTs, and captures the sign of the response consistently for ENF. It is useful to also compare our theory, the observed uncertainty in the ET response, and PET, as PET is often used as a proxy for ET in hydrometeorological studies and drought indices. Again, as in Section \ref{sign_term}, the PET sensitivity to VPD will be constant at about 3.5. In Figure \ref{agu_fig} the ET response to shifts in dryness is always far smaller than 3.5 in the observations, even with all uncertainty included. So the PET response to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budget and dryness response at the land surface, such as used in several drought indices. However our theory (Equation \ref{et}) could be used with published $uWUE$ and $g1$ data, as an alternative to PET to estimating land response and dryness, and would more accurately account for the response to atmospheric drying.

@@

Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty ($\sigma$). Each column represents a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. For any observations to the left of this line $\frac{\partial \; ET}{\partial \; VPD}$ should be negative (plant response dominates), and any observations the right of the line should be positive (atmospheric demand dominates). In addition to seeing how our theoretically-derived VPD threshold, $VPD_{crit}$, matches observations, these analysis are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

Figure \ref{real} generally confirms expectations from Figure \ref{agu_fig}. By comparing different columns of Figure \ref{real} we can evaluate the impact of $\sigma$ and aerodynamic conductance ($g_a$) on the variability of ET response ($\frac{\partial \; ET}{\partial \; VPD}$). Aerodynamic conductance's scaling (included in columns 1 and 3) alters the magnitude considerably, particularly between PFTs. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$, as observed in Figure \ref{agu_fig}. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is held constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, where the full complexity plots (Columns 1 and 2) do not match well with $\sigma$ held fixed (Columns 3 and 4).

@@

In general Figures \ref{agu_fig} and \ref{real} match our expectations. Our $\sigma$-based method of uncertainty propagation blurs the idealized expectations the most for GRA and CRO. We therefore have the most confidence in our conclusion based on Equation \ref{d_et} for PFTs CSH, DBF, and ENF, as the sign of the derivative with uncertainty included matches (to leading order) the story when $\sigma$ is held fixed.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./agu_fig.pdf}
\caption{Comparison of the sign term with model uncertainty included (box plots) to the sign term as calculated with simplifying assumptions (blue line).}
\label{agu_fig}
\end{figure}

\begin{figure}[!p]
\centering
\centerline{\includegraphics[width=1.4\textwidth]{./fig06.png}}
\caption{Scatter plots of $\frac{\partial \; ET}{\partial \; VPD}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, as labeled: Column 1: $\frac{\partial \; ET}{\partial \; VPD}$; Column 2: $\frac{\partial \; ET}{\partial \; VPD}$ normalized by $g_a$; Column 3: $\frac{\partial \; ET}{\partial \; VPD}$ with $\sigma$ held fixed at 1; and Column 4: $\frac{\partial \; ET}{\partial \; VPD}$ normalized by $g_a$ and with $\sigma$ held fixed. Please note differences in the colorbar scale.}
\label{real}
\end{figure}

@@

Indeed, for all PFTs, there is some slight dependence of $\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure \ref{swc}). The portion of observations for which our theory is biased by SWC-limitations varies by PFT, due to the nonlinear threshold at which soil moisture availability limits plant function. For CRO and DBF, soil water content only matters for about the lowest 5\% of observations (each box is 5\% of observations in Figure \ref{swc}). And, CRO, relative to DBF, has a very weak dependence on soil moisture, which is reflective of the high likelihood that CRO sites are optimally irrigated and not water stressed. For ENF, systematic SWC-induced biases in $\sigma \cdot uWUE$ emerge in about the lowest 30\% of SWC conditions. In crontast to ENF, CRO, and DBF, for GRA the relationship between $\sigma \cdot uWUE$ and SWC is more linear and affects a greater portion of observations; about 60\% of observations. Clearly, for GRA soil water frequently impacts plant function and alters ET response, and our theory is limited. It is therefore not suprising that our theory tested poorly againt the data for GRA, relative to ENF, DBF, and GRA (Section \ref{testing}). For all other PFTs occurence of soil mosture impacts was rare enough to not manifest itself in bulk statistics and plots\footnote{note that interpreting the soil moisture dependence for CSH is difficult as their are only two sites for that PFT. So systematic soil moisture relationshps are not apparent in Figure \ref{swc}, but instead blurred by the superposition of the two sites. We argue taht the soil moisture dependence is still there, but we don't have enough statistcs to comment on it}.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./swc_boxplot.pdf}
\caption{$\sigma \cot uWUE$ as a function of PFT and SWC. Each boxplot represents 5\% of all observations.}
\label{swc}
\end{figure}

>>>>

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./map.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  a placeholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

@@

@@

By construction, the variability in the $\sigma$ term (Equation \ref{sigma}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be close to 1. Figure \ref{joint_vpd_sigma} presents the histogram of calculated $\sigma$s with outliers removed (lowest and highest 5\% percent). All remaining $\sigma$ values are close to unity (i.e. $O(1)$), which provides further confidence in the model framework and in the data quality.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./joint_vpd_sigma.pdf}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$.}
\label{joint_vpd_sigma}
\end{figure}

An additional concern is that $\sigma$ may in fact be correlated with $VPD$, in which case the dependence would need to be accounted for when taking the derivative. However, the joint distribution of $\sigma$ and VPD shows that $\sigma$ is very weakly a function of VPD (Figure \ref{joint_vpd_sigma}). Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $VPD$.

@@

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive CRO and water conservative CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$. It also has a slightly higher uWUE ($2.32$ -units-) compared to CSH ($uWUE=2.11$ -units-). \explain{MAYBE A DUMB QUESTION BUT CAN'T uWUE AND G1 BE ANALYTICALLY RELTED> I THINK AT LESAT YOU SHOULD MAKE CLEAR THAT UWUE IS AT THE ECOYSTEM SCALE (DID YOU FIT IT OR USED PUBLISHED DATA-I AM NOW CONFUSED)}. These differences in slope parameter cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always increases with increasing VPD for the more water intensive CRO, while ET almost always decreases with increasing VPD for the more water conservative CSH (Figure \ref{idealized_sign}). CRO evolved or were bred to prioritize GPP and yield and are thus not water conservative. They are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./idealized_sign.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***I think I'm going to remove the bottom plot: its discussion didn't fit in well with the text, and the box plots test inclusion of uncertainty***}}
\label{idealized_sign}
\end{figure}

As expected, the slope parameter ($g_1$) is a primary determinant of the VPD dependence for the sign term shown in Figure \ref{idealized_sign}. Plants that are more conservative (small $g_1$) will tend to reduce ET with increasing VPD, and will be very effective at reducing ET, especially at low VPD. However, at very high VPD, gradients in vapor pressure at the leaf scale will be so strong that they will begin to dominate leaf-scale plant strategies of stomatal (partial) closure (parameterized with $g_1$). As a result, ET response will begin to asymptote towards its ecosystem-scale values irregardless of leaf-scale strategy and stomatal regulation (parameterized with $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf stomatal response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{idealized_sign}.

To summarize our theoretical insights, Figure \ref{idealized_sign} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET with increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD response dependence, while CRO ($g_1 = 183.1$) has the smallest VPD dependence. ENF are less willing to trade water for GPP, so the difference between the response at low VPD when leaf response dominates and high VPD when vapor pressure gradients dominate is much greater. CROs on the other hand have a more aggressive water use strategy and are more willing to trade water for photosynthesis, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively small.

Figure \ref{idealized_sign} clearly shows, according to our theory, that for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to reduce ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates in typical environmental conditions, emphasizing the water conservative strategy of those plants. For CRO on the other hand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants operate with an aggressive water usage strategy, are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on environmental conditions, with about half of the observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, ET response to VPD is still far smaller than it would be for potential evaporation $\partial PET/\partial VPD$, i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface response to a change in VPD will always me more negative than if the plants are not included in the physics (PET).

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./idealized_scale.pdf}
\caption{Primary sources of variability for the scaling term, as a function of PFT. Note the change in axis scale for each plot. The 5th-95th percentile range of temperature is plotted at the 5th, 25th, 50th, 75th, and 95th percentiles of aerodynamic conductance, as observed for each PFT's sites.}
\label{scale_vary}
\end{figure}

@@

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
While Section \ref{stats_sec} presented simple statistics of observed response, we now compare more sophisticated distributions of the observed response to our theory presented in Section \ref{sign_term}. Figure \ref{test_sign} shows observed the distribution of of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./test_sign.pdf}
\caption{Comparison of the sign term with model uncertainty included (box plots) to the sign term as calculated with simplifying assumptions (blue line).}
\label{test_sign}
\end{figure}

@@

As Section \ref{stats_sec} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, due to some bias in $uWUE$. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics of Section \ref{stats_sec}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance tend to not only observe a single species but many differnent ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{test_sign}.

@@

While the above discussion shows that our theory has some limitations when applied to some PFTs, especially for crops and grasslands, it does well for DBF and CSH PFTs, and captures the sign of the response consistently for ENF. It is useful to also compare our theory, the observed uncertainty in the ET response, and PET, as PET is often used as a proxy for ET in hydrometeorological studies and drought indices. Again, as in Section \ref{sign_term}, the PET sensitivity to VPD will be constant at about 3.5. In Figure \ref{test_sign} the ET response to shifts in dryness is always far smaller than 3.5 in the observations, even with all uncertainty included. So the PET response to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budget and dryness response at the land surface, such as used in several drought indices. However our theory (Equation \ref{et}) could be used with published $uWUE$ and $g1$ data, as an alternative to PET to estimating land response and dryness, and would more accurately account for the response to atmospheric drying.

@@

Figure \ref{data_scatter} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty ($\sigma$). Each column represents a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. For any observations to the left of this line $\frac{\partial \; ET}{\partial \; VPD}$ should be negative (plant response dominates), and any observations the right of the line should be positive (atmospheric demand dominates). In addition to seeing how our theoretically-derived VPD threshold, $VPD_{crit}$, matches observations, these analysis are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

\begin{figure}[!p]
\centering
\centerline{\includegraphics[width=1.4\textwidth]{./data_scatter.png}}
\caption{Scatter plots of $\frac{\partial \; ET}{\partial \; VPD}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, as labeled: Column 1: $\frac{\partial \; ET}{\partial \; VPD}$; Column 2: $\frac{\partial \; ET}{\partial \; VPD}$ normalized by $g_a$; Column 3: $\frac{\partial \; ET}{\partial \; VPD}$ with $\sigma$ held fixed at 1; and Column 4: $\frac{\partial \; ET}{\partial \; VPD}$ normalized by $g_a$ and with $\sigma$ held fixed. Please note differences in the colorbar scale.}
\label{data_scatter}
\end{figure}

Figure \ref{data_scatter} generally confirms expectations from Figure \ref{test_sign}. By comparing different columns of Figure \ref{data_scatter} we can evaluate the impact of $\sigma$ and aerodynamic conductance ($g_a$) on the variability of ET response ($\frac{\partial \; ET}{\partial \; VPD}$). Aerodynamic conductance's scaling (included in columns 1 and 3) alters the magnitude considerably, particularly between PFTs. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$, as observed in Figure \ref{test_sign}. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is held constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, where the full complexity plots (Columns 1 and 2) do not match well with $\sigma$ held fixed (Columns 3 and 4).

@@

In general Figures \ref{test_sign} and \ref{data_scatter} match our expectations. Our $\sigma$-based method of uncertainty propagation blurs the idealized expectations the most for GRA and CRO. We therefore have the most confidence in our conclusion based on Equation \ref{d_et} for PFTs CSH, DBF, and ENF, as the sign of the derivative with uncertainty included matches (to leading order) the story when $\sigma$ is held fixed.

@@

Indeed, for all PFTs, there is some slight dependence of $\sigma \cdot uWUE$ on SWC, especially at low SWC (Figure \ref{swc_boxplot}). The portion of observations for which our theory is biased by SWC-limitations varies by PFT, due to the nonlinear threshold at which soil moisture availability limits plant function. For CRO and DBF, soil water content only matters for about the lowest 5\% of observations (each box is 5\% of observations in Figure \ref{swc_boxplot}). And, CRO, relative to DBF, has a very weak dependence on soil moisture, which is reflective of the high likelihood that CRO sites are optimally irrigated and not water stressed. For ENF, systematic SWC-induced biases in $\sigma \cdot uWUE$ emerge in about the lowest 30\% of SWC conditions. In crontast to ENF, CRO, and DBF, for GRA the relationship between $\sigma \cdot uWUE$ and SWC is more linear and affects a greater portion of observations; about 60\% of observations. Clearly, for GRA soil water frequently impacts plant function and alters ET response, and our theory is limited. It is therefore not suprising that our theory tested poorly againt the data for GRA, relative to ENF, DBF, and GRA (Section \ref{testing}). For all other PFTs occurence of soil mosture impacts was rare enough to not manifest itself in bulk statistics and plots\footnote{note that interpreting the soil moisture dependence for CSH is difficult as their are only two sites for that PFT. So systematic soil moisture relationshps are not apparent in Figure \ref{swc_boxplot}, but instead blurred by the superposition of the two sites. We argue taht the soil moisture dependence is still there, but we don't have enough statistcs to comment on it}.

\begin{figure}[!p]
\centering
\includegraphics[width=\textwidth]{./swc_boxplot.pdf}
\caption{$\sigma \cot uWUE$ as a function of PFT and SWC. Each boxplot represents 5\% of all observations.}
\label{swc_boxplot}
\end{figure}

@@
